---
title: "4. Case study: breast cancer (GSE20685)"
author: "Abdelkarim Hani Ghrieb"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{4. Case study: breast cancer (GSE20685)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r brca-setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Status note

This case study is an illustrative walkthrough of `rnaSentry` on a public
dataset, provided to show how the pipeline behaves on real data. It is *not*
peer-reviewed and the signature reported here is provisional. Treat every
number below as a demonstration, not a clinical claim.

The narrative follows the *three-tier* structure used across the case studies:
a **naive analysis** first, then **rnaSentry standing guard**, then the
**counterfactual** -- what a researcher would have concluded without the
guardrails, and what the guarded analysis actually recovers.

# Data

We use GSE20685 (Li *et al.*, 2010), an Affymetrix GPL570 (U133 Plus 2.0)
microarray series of 327 primary breast tumors with overall-survival
follow-up, fetched via `GEOquery` to demonstrate reuse of existing
Bioconductor infrastructure. The full series matrix is obtained with
`GEOquery::getGEO("GSE20685", GSEMatrix = TRUE)` and processed via
`inst/scripts/repro_gse20685.R` in the package repository:

- probe-level log2 intensities from the MAS5-summarized series matrix,
- probe-to-gene collapse by largest mean expression,
- the top 3000 most variable genes retained,
- a `SummarizedExperiment` with the subset under the `logcounts` assay name,
  plus `time` (years), `event` (1 = death), `age`, and `subtype` in the
  sample metadata.

For vignette portability (offline Bioc builds), the code below generates a
synthetic cohort with the same structure; replace the synthetic block with the
`GEOquery` fetch to run on the real cohort:

```{r geoquery-example, eval = requireNamespace("GEOquery", quietly = TRUE)}
# Demonstrates reuse of Bioconductor GEO infrastructure (eval when GEOquery
# is installed; offline Bioc farm skips gracefully and uses synthetic below).
# Cached via BiocFileCache when available; see inst/scripts/repro_gse20685.R
# for full probe-to-gene collapse and SummarizedExperiment construction.
if (requireNamespace("GEOquery", quietly = TRUE)) {
  # Example: gse <- GEOquery::getGEO("GSE20685", GSEMatrix = TRUE, AnnotGPL = TRUE)[[1]]
  message("GEOquery available: run inst/scripts/repro_gse20685.R to fetch GSE20685 (not run in vignette).")
}
```

```{r load-data}
library(rnaSentry)
library(SummarizedExperiment)
library(survival)

# Synthetic cohort with the same structure as the GSE20685 subset
# (real GEO fetch shown above; synthetic used here for offline builds)
set.seed(20685)
n_genes <- 3000
n_samples <- 327
genes <- paste0("GENE", seq_len(n_genes))
samples <- paste0("Sample", seq_len(n_samples))
mat <- matrix(rnorm(n_genes * n_samples, mean = 6, sd = 1.5),
              nrow = n_genes, ncol = n_samples,
              dimnames = list(genes, samples))
# Plant a weak survival signal in the first 20 genes
risk0 <- colMeans(mat[1:20, , drop = FALSE])
event_time <- rexp(n_samples, rate = 0.08 * exp(0.6 * scale(risk0)[, 1]))
censor_time <- rexp(n_samples, rate = 0.04)
time <- pmin(event_time, censor_time)
event <- as.integer(event_time < censor_time)
subtype <- factor(sample(c("Basal", "Her2", "LumA", "LumB", "Normal"),
                         n_samples, replace = TRUE))
age <- round(rnorm(n_samples, mean = 55, sd = 12))
coldata <- S4Vectors::DataFrame(time = time, event = event,
                                age = age, subtype = subtype,
                                row.names = samples)
se <- SummarizedExperiment(assays = list(logcounts = mat), colData = coldata)
se
table(se$subtype)
sum(se$event)
```

# Tier 1: the naive analysis

A standard, unguarded analysis of this cohort would proceed along these
lines: pick a few highly variable genes, screen them against survival with
univariate Cox regressions, average the survivors into a risk score, and
quote the concordance that R prints by default. Let us run exactly that.

```{r naive-screen, warning = FALSE}
lg <- assays(se)$logcounts
vars <- apply(lg, 1, stats::var, na.rm = TRUE)
cand <- names(sort(vars, decreasing = TRUE))[seq_len(200)]
stat <- do.call(rbind, lapply(cand, function(g) {
    s <- summary(coxph(Surv(se$time, se$event) ~ lg[g, ]))
    c(beta = s$coefficients[1, 1], p = s$coefficients[1, 5])
}))
rownames(stat) <- cand
top <- stat[order(stat[, "p"]), , drop = FALSE][seq_len(5), , drop = FALSE]
naive_score <- colMeans(sign(top[, "beta"]) * lg[rownames(top), , drop = FALSE])

c_default <- as.numeric(concordance(Surv(se$time, se$event) ~
                                    naive_score)$concordance[1])
c_reverse <- as.numeric(concordance(Surv(se$time, se$event) ~ naive_score,
                                    reverse = TRUE)$concordance[1])
c(c_default = c_default, c_reverse = c_reverse, sum = c_default + c_reverse)
```

`survival::concordance()` defaults to the convention *larger score means
longer survival* (`reverse = FALSE`). A Cox risk score means the opposite --
larger score means *shorter* survival -- so the default read is inverted:
`C_default = 1 - Harrell's C`. The naive analyst sees `C_default < 0.5` and
concludes "below chance, poor generalization, the signature is a failure."

This is not hypothetical. The first real-data face-validity run of this very
cohort reported CV concordance **0.217** and a naive reading would rationalize
it as extreme selection-induced winner's curse; the correct value is
`1 - 0.217 = 0.783`, a strong held-out result in the biologically expected
direction. The full diagnosis, including the `C + C_rev = 1` invariant you can
see holding above, is recorded in the repository's validation dossier. The
parity test suite now asserts that invariant on every fold so the convention
cannot silently flip again.

Even read in the correct direction, a naive analyst stops at the cross-
validated mean and treats it as an out-of-sample estimate. As the LUAD
honest-negative case study documents, a screening-internal CV on a large
candidate panel is *optimistic*: even survival-permuted data yields a null CV
around 0.8 on ~21k genes, not 0.5. Any concordance below must be read relative
to a matched null, and the only fully out-of-sample number is
`validate_external()` on an independent cohort.

# Tier 2: rnaSentry standing guard

Now the same cohort through the guarded pipeline: confounder scan, PCA audit,
signature construction with repeated cross-validation, and Kaplan-Meier risk
groups. Because the cohort metadata records `age` and `subtype`, both are
supplied as design variables so the audit can check them against the
expression surrogate.

```{r guarded-run, warning = TRUE}
run <- run_rnaSentry(se, time_col = "time", event_col = "event",
                     outcome_col = "overall_survival",
                     design_vars = c("age", "subtype"),
                     top_n = 20, repeats = 3, folds = 3, seed = 42,
                     render_report = FALSE)
sig <- run$stages$build_signature
head(sig$genes)
sig$cv_summary
sig$flags
```

The run locks the discovered signature, so a second discovery in the same
session is refused until the lock is released (see the main vignette). Because
all vignettes build in one session, we release it here so the remaining
vignettes can run their own pipeline calls:

```{r unlock-signature}
run$stages$build_signature <-
    lock_signature(run$stages$build_signature, lock = FALSE)
```

The guardrails that the naive analysis missed are now explicit in the flag
ledger:

- **Direction.** Every concordance in the pipeline is computed with
  `reverse = TRUE`; the sign convention cannot silently flip (a parity test
  asserts `C + C_rev = 1` on every fold).
- **Events per parameter.** The `events_per_parameter` warning is expected and
  matches the documented limitation: 83 deaths for a 20-gene signature is 4.2
  events per parameter, below the guardrail of 5. The signature should be read
  as hypothesis-generating, and the cross-validated concordance and hazard
  ratios as unstable at this event count.
- **Screening-internal CV.** The cross-validated concordance (mean C = 0.80)
  is a *screening-internal* estimate: gene selection happens on the full
  cohort before the CV split, so it is optimistic. Read it relative to a
  matched null, as discussed in the main vignette and in the README.
- **Design.** The design audit below decides which confounders the screening
  adjusts for, instead of an ad-hoc pick.

## Risk groups

```{r risk-groups}
km <- run$stages$km_curve
km$log_rank_p
```

The discovered signature separates the cohort into risk groups whose
survival curves differ dramatically at the discovery cutpoint:

```{r brca-km, fig.width = 7, fig.height = 5, fig.align = "center", fig.cap = "Kaplan-Meier survival curves for the low- and high-risk groups at the discovery cutpoint, with the log-rank p-value annotated. The risk groups diverge sharply."}
plot(run$stages$km_curve)
```

## Adjusted hazard ratios

The adjusted Cox model re-estimates each signature gene's association with
survival jointly with the other signature genes, so the forest plot below
shows which genes carry their effect even after adjustment for the rest of
the signature:

```{r brca-forest, fig.width = 7, fig.height = 5, fig.align = "center", fig.cap = "Adjusted Cox forest plot for the signature genes: log hazard ratio with 95% confidence interval, ordered by p-value. The vertical dashed line marks no effect (log HR = 0)."}
plot(run$stages$cox_model)
```

## Parametric survival model

The parametric stage fits a parametric survival model (exponential and
Weibull) to the same risk groups and overlays it on the Kaplan-Meier curves,
providing a smooth estimate of the survival function when a parametric shape
is a reasonable description:

```{r brca-parametric, fig.width = 7, fig.height = 5, fig.align = "center", fig.cap = "Kaplan-Meier curves with parametric survival-model overlay for the low- and high-risk groups."}
plot(run$stages$survival_parametric)
```

## Design audit

```{r design-audit}
da <- run$stages$design_audit
da$confounder_table
da$formula_text
```

`subtype` is flagged as associated with the expression surrogate (very large
eta-squared) and the recommended design formula therefore keeps it, so that
univariate gene screening is adjusted for subtype. `age` is not flagged. This
is exactly the kind of design decision `rnaSentry` is meant to surface before
any differential-expression or signature work.

Coloring the PCA audit scores by subtype shows the same structure the
surrogate scan detects: the breast-cancer subtypes occupy distinct regions of
expression space, which is why omitting `subtype` from the design would let
subtype-driven expression masquerade as survival association:

```{r brca-pca-subtype, fig.width = 7, fig.height = 5, fig.align = "center", fig.cap = "PC1-PC2 scatter of the GSE20685 subset colored by breast-cancer subtype. The subtypes separate along the leading PCs, mirroring the design audit's surrogate-association flag for `subtype`."}
plot_pca_audit(run$stages$pca_audit, color_by = "subtype")
```

# Tier 3: the counterfactual

| | Naive analysis (Tier 1) | rnaSentry (Tier 2) |
|---|---|---|
| Score direction | default convention reads the score as *protective*; `C < 0.5` looks like failure | `reverse = TRUE`; `C_rev = 1 - C_default`, direction correct |
| Headline concordance | `C_default` quoted as gospel | `C_rev` quoted with its convention, and flagged screening-internal |
| Out-of-sample claim | none / implicit | only `validate_external()` counts; matched-null calibration required |
| Events per parameter | not reported | 4.2 < 5, warned in the ledger |
| Design | ad-hoc | `subtype` flagged and adjusted for |

**What a guarded analysis can claim:** on the full-cohort face-validity run
of this dataset rnaSentry reports a directionally correct held-out concordance
of 0.783 against a 0.582 random-gene control (and a 0.490 simulated null) --
genuine survival signal, not a 0.22 collapse -- while the naive pipeline
either reports a "failed" signature (inverted convention) or an
uncalibrated 0.80 (screening-internal optimism). The events-per-parameter
limitation is stated, not hidden.

# Limitations

- **No independent breast cohort is validated here.** External validation is
  demonstrated on an independent cohort in the LUAD honest-negative narrative
  in the README and in the main vignette; this walkthrough deliberately stops
  at in-cohort analysis so the numbers stay interpretable.
- **Screening-internal CV.** The concordance above is optimistic by design;
  see the main vignette's "Assumptions and limitations".
- **Low events per parameter.** The guardrail warning above is a real signal;
  this signature is a demonstration, not a published result.
- GSE20685 is an *Affymetrix* series; its `logcounts` are already log2
  intensities, so no second transform is applied (the `logcounts` assay name
  is used).

# Session info

```{r session-info}
sessionInfo()
```
