---
title: "Case study: a breast cancer prognostic signature (GSE20685)"
author: "Abdelkarim Hani Ghrieb"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Case study: breast cancer (GSE20685)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Status note

This case study is an illustrative walkthrough of `rnaSentry` on a public
dataset, provided to show how the pipeline behaves on real data. It is *not*
peer-reviewed and the signature reported here is provisional until a preprint
is published. Treat every number below as a demonstration, not a clinical
claim.

# Data

We use a bundled subset of GSE20685 (Li *et al.*, 2010), an Affymetrix
GPL570 (U133 Plus 2.0) microarray series of 327 primary breast tumors with
overall-survival follow-up. The series matrix was downloaded from GEO and
processed in `validation/repro_gse20685.R` (dev-only script, not shipped):

- probe-level log2 intensities from the MAS5-summarized series matrix,
- probe-to-gene collapse by largest mean expression,
- the top 3000 most variable genes retained for the bundled subset,
- a `SummarizedExperiment` with the subset under the `logcounts` assay name,
  plus `time` (years), `event` (1 = death), `age`, and `subtype` in the
  sample metadata.

```{r}
library(rnaSentry)
library(SummarizedExperiment)

se <- readRDS(system.file("extdata", "gse20685_case_study.rds",
                          package = "rnaSentry"))
se
table(se$subtype)
sum(se$event)
```

# Running the pipeline

We run the full pipeline in one call: confounder scan, PCA audit, signature
construction with repeated cross-validation, and Kaplan-Meier risk groups.
Because the cohort metadata records `age` and `subtype`, both are supplied as
design variables so the audit can check them against the expression
surrogate.

```{r, warning = TRUE}
run <- run_rnaSentry(se, time_col = "time", event_col = "event",
                     outcome_col = "overall_survival",
                     design_vars = c("age", "subtype"),
                     top_n = 20, repeats = 3, folds = 3, seed = 42,
                     render_report = FALSE)
sig <- run$stages$build_signature
head(sig$genes)
sig$cv_summary
sig$flags
```

The run locks the discovered signature, so a second discovery in the same
session is refused until the lock is released (see the main vignette). Because
all vignettes build in one session, we release it here so the remaining
vignettes can run their own pipeline calls:

```{r}
run$stages$build_signature <-
    lock_signature(run$stages$build_signature, lock = FALSE)
```

The `events_per_parameter` warning is expected and matches the documented
limitation: 83 deaths for a 20-gene signature is 4.2 events per parameter,
below the guardrail of 5. The signature should be read as hypothesis-generating, and the cross-validated concordance and hazard ratios as
unstable at this event count.

The cross-validated concordance (mean C = 0.80) is a *screening-internal*
estimate: gene selection happens on the full cohort before the CV split, so
it is optimistic. Read it relative to a matched null, as discussed in the
main vignette and in the README.

# Risk groups

```{r}
km <- run$stages$km_curve
km$log_rank_p
```

The discovered signature separates the cohort into risk groups whose
survival curves differ dramatically at the discovery cutpoint.

# Design audit

```{r}
da <- run$stages$design_audit
da$confounder_table
da$formula_text
```

`subtype` is flagged as associated with the expression surrogate (very large
eta-squared) and the recommended design formula therefore keeps it, so that
univariate gene screening is adjusted for subtype. `age` is not flagged. This
is exactly the kind of design decision `rnaSentry` is meant to surface before
any differential-expression or signature work.

# Limitations

- **No independent breast cohort is validated here.** External validation is
  demonstrated on an independent cohort in the LUAD honest-negative narrative
  in the README and in the main vignette; this walkthrough deliberately stops
  at in-cohort analysis so the numbers stay interpretable.
- **Screening-internal CV.** The concordance above is optimistic by design;
  see the main vignette's "Assumptions and limitations".
- **Low events per parameter.** The guardrail warning above is a real signal;
  this signature is a demonstration, not a published result.
- GSE20685 is an *Affymetrix* series; its `logcounts` are already log2
  intensities, so no second transform is applied (the `logcounts` assay name
  is used).

# Session info

```{r}
sessionInfo()
```
