---
title: "Case study: why small cohorts fail silently"
author: "Abdelkarim Hani Ghrieb"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Case study: small cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Status note

This is a synthetic-data walkthrough demonstrating how the
`events_per_parameter` guardrail and fold-level reporting expose unstable
signatures. It is a vignette, not a study.

# Setup

We build a 30-sample cohort with 23 events and a real (if weak) survival
signal driven by five genes, then ask for a 5-gene signature. This is 4.6
events per parameter -- below the guardrail of 5.

```{r, warning = TRUE}
library(rnaSentry)
library(SummarizedExperiment)

set.seed(30)
n_sm <- 30
counts <- matrix(rpois(60 * n_sm, lambda = 500), nrow = 60, ncol = n_sm,
                 dimnames = list(paste0("gene", 1:60), paste0("S", 1:n_sm)))
sig_expr <- colMeans(counts[1:5, , drop = FALSE])
risk <- scale(sig_expr)[, 1] * 0.5
event_time <- rexp(n_sm, rate = 0.05 * exp(0.9 * risk))
censor_time <- rexp(n_sm, rate = 0.02)
cd <- DataFrame(time = pmin(event_time, censor_time),
                event = as.integer(event_time < censor_time),
                row.names = colnames(counts))
se <- SummarizedExperiment(assays = list(counts = counts), colData = cd)

sig <- build_signature(se, "time", "event", top_n = 5,
                       repeats = 5, folds = 5, seed = 30)
```

# The guardrail fires

```{r}
sig$flags
```

The `events_per_parameter` warning fires immediately, and is also recorded in
the stage's flag ledger so it propagates into the rendered report.

# The mean hides the spread

```{r}
sig$cv_summary
```

The mean cross-validated concordance is 0.74 -- which a careless reader would
call "decent". Look at the fold-level table instead:

```{r}
sig$cv_results
```

The per-fold concordances range from **0.50 to 1.00** across repeats and
folds. With only a handful of held-out samples per fold, single-fold C
estimates are nearly coin flips, and the standard deviation (0.14) is
enormous relative to the mean. A signature like this is unstable and should
be treated as hypothesis-generating only.

# Reading

The takeaway is exactly the documented "small-cohort trap": with fewer than
~5 events per signature gene, neither the concordance nor the hazard ratios
are trustworthy, and a *mean* C-index hides the instability that the
fold-level table exposes. This is why `build_signature()` reports the
per-fold distribution, not just the mean, and why the guardrail warns rather
than silently proceeding.

# Session info

```{r}
sessionInfo()
```
