---
title: "Case study: the confounder audit on a batch-confounded cohort"
author: "Abdelkarim Hani Ghrieb"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Case study: confounder audit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Status note

This is a synthetic-data walkthrough used to verify `design_audit()`
behavior on a deliberately confounded cohort. It is a vignette, not a study.

# Setup

We build a cohort of 180 samples with two design variables that are almost
interchangeable:

- `region` is a 3-level factor derived by cutting a latent normal variable;
- `batch` is a *noisy copy* of `region` (80% of samples keep their `region`
  label; the rest are shuffled).

We also plant a real batch effect in expression: six genes are multiplied by
a batch-dependent factor. The surrogate expression PC is therefore driven by
`batch`, and `batch`/`region` are nearly collinear.

```{r}
library(rnaSentry)
library(SummarizedExperiment)

set.seed(12)
n <- 180
latent <- rnorm(n)
region <- factor(cut(latent, breaks = 3, labels = c("R1", "R2", "R3")))
batch <- as.character(region)
for (i in seq_len(n))
    if (runif(1) < 0.2) batch[i] <- sample(c("R1", "R2", "R3"), 1)
batch <- factor(batch)

counts <- matrix(rpois(50 * n, lambda = 200), nrow = 50, ncol = n,
                 dimnames = list(paste0("g", 1:50), paste0("S", 1:n)))
for (b in 1:3) {
    idx <- which(batch == levels(batch)[b])
    counts[1:6, idx] <- counts[1:6, idx] * (1 + 0.6 * b)
}

cd <- DataFrame(batch = batch, region = region, row.names = colnames(counts))
se <- SummarizedExperiment(assays = list(counts = counts), colData = cd)
```

# The audit

```{r}
da <- design_audit(se, design_vars = c("batch", "region"),
                   redundant_effect_size = 0.5)
```

## Pairwise redundancy

```{r}
da$pairwise_table
```

`batch` and `region` are almost interchangeable (Cramer's V = 0.82), and the
pair is flagged `redundant`. Including both in a model would double-count
nearly the same information.

## Confounder scan

```{r}
da$confounder_table
```

Both variables are flagged as associated with the expression surrogate
`batch` with eta-squared 0.97, `region` with epsilon-squared 0.55 (it leaks
the same signal through its overlap with `batch`).

## Recovered design

```{r}
da$formula_text
```

The audit therefore recommends the minimal formula `~ batch`, dropping the
redundant `region` term. This is the behavior verified by the test suite
(`confounder-detection` and `design-audit` blocks): a redundant design term
is reported, not silently kept.

# Reading

The lesson is that the recommended formula is only as good as the design
variables you supply. `design_audit()` will not invent a missing confounder;
it audits what it is given. Supply every variable that could plausibly
confound expression (batch, plate, sex, subtype), then trust the recommended
formula over an ad-hoc pick.

# Session info

```{r}
sessionInfo()
```
