---
title: "Introduction to rnaSentry"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Introduction to rnaSentry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`rnaSentry` automates the statistical pipeline used to go from a raw bulk
RNA-seq count matrix to a validated prognostic gene signature: quality
control and sample-identity auditing, confounder detection prior to
differential expression, signature construction with cross-validation,
survival modeling, and external validation. Every automated decision is
reported with the statistic that justified it.

This vignette demonstrates the pipeline's first stage: quality control and
biological-sex verification.

# Installation

```{r, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("rnaSentry")
```

# Quality control

`qc_explore()` takes a `SummarizedExperiment` of raw counts and audits it
for duplicate sample identifiers, missing metadata, non-integer count
values, and library-size outliers.

```{r}
library(rnaSentry)
library(SummarizedExperiment)

set.seed(1)
counts <- matrix(rpois(600, lambda = 200), nrow = 60, ncol = 10,
                  dimnames = list(paste0("gene", 1:60), paste0("S", 1:10)))
counts[, 1] <- counts[, 1] * 50L
coldata <- DataFrame(condition = rep(c("tumor", "normal"), each = 5),
                      row.names = paste0("S", 1:10))
se <- SummarizedExperiment(assays = list(counts = counts), colData = coldata)

qc <- qc_explore(se)
qc
```

# Biological sex verification

`sex_check()` compares reported sex in sample metadata against sex inferred
from XIST and Y-chromosome marker gene expression, to catch sample swaps
or metadata errors before they propagate downstream.

```{r}
genes <- c("XIST", "RPS4Y1", "DDX3Y", "KDM5D")
expr <- matrix(c(950, 5, 5, 5,   # sample looks female
                  5, 400, 380, 410),  # sample looks male
                nrow = 4, dimnames = list(genes, c("S1", "S2")))
coldata2 <- DataFrame(sex = c("F", "F"), row.names = c("S1", "S2")) # S2 mislabeled
se2 <- SummarizedExperiment(assays = list(counts = expr), colData = coldata2)

sex_check(se2)
```

# Session info

```{r}
sessionInfo()
```
