---
title: "rnaSentry analysis report"
author: "rnaSentry"
date: "`r Sys.Date()`"
output: html_document
params:
  stages: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Pipeline summary

```{r summary, results="asis"}
cat(sprintf("Stages run: %s.\n",
            paste(names(params$stages), collapse = ", ")))
```

## Pipeline assumptions and scope

This report was generated by `rnaSentry`, which models standard
right-censored overall survival from bulk RNA-seq cohorts using a linear Cox
proportional-hazards risk score. The pipeline assumes:

- **Input data.** Bulk RNA-seq with gene symbols as rownames and either raw
  integer counts (which are log2-transformed internally) or an assay already
  named `logcounts`/`vst`. Pre-transformed data in a `counts`-named slot is
  flagged as `possibly_log_scaled` because it would be double-logged.
- **Design.** A single prognosis end point recorded as right-censored
  (time, event) columns. Competing risks, time-varying covariates, and left
  truncation are not modeled.
- **Score.** The signature is a weighted linear combination of gene
  expression. Higher scores indicate higher hazard (shorter survival), and
  all concordance estimates use this convention.
- **Selection.** Signature genes are selected by univariate Cox screening
  and a joint multivariate Cox fit. The pipeline reports, but does not
  correct, batch or design effects.

Interpret results cautiously when a stage flagged low event counts per
parameter (`events_per_parameter`), sparse risk-group events
(`sparse_events`), or a potential proportional-hazards violation
(`proportional_hazards`). See the audit trail below for every issue raised.

## Audit trail

```{r audit, results="asis"}
flag_tables <- lapply(names(params$stages), function(nm) {
  f <- params$stages[[nm]]$flags
  if (is.null(f) || !is.data.frame(f) || nrow(f) == 0) return(NULL)
  if (!"stage" %in% colnames(f)) f$stage <- nm
  f
})
flag_tables <- Filter(function(x) !is.null(x), flag_tables)
if (length(flag_tables) > 0) {
  all_flags <- do.call(rbind, flag_tables)
  knitr::kable(all_flags)
} else {
  cat("No issues flagged at any stage.")
}
```

## Stage results

```{r stages, results="asis"}
for (nm in names(params$stages)) {
  cat("\n\n###", nm, "\n\n")
  print(params$stages[[nm]])
}
```
